---
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 2
---

# grattan

Australian Tax Policy Analysis

# Overview
```{r eval=FALSE}
install.packages("grattan")
```

```{r loadPackages}
library(grattan)
```

## `income_tax`
Calculates the income tax for a given taxable income and financial year:

```{r income_tax_1}
income_tax(50e3, "2015-16")
```

### With sample files
`income_tax` is designed to work well with the ATO's sample files. You can obtain the sample files from my repo:

```{r install_taxstats}
# install.packages("taxstats", repos = "https://hughparsonage.github.io/tax-drat")
library(taxstats)

library(data.table) 
library(magrittr)
```

Simply pass the sample file to `.dots.ATO` and the complexities of things like Medicare levy and the Seniors and Pensioners Tax Offset are handled for you. For example:

```{r income_tax_s1314}
s1314 <- as.data.table(sample_file_1314)
s1314 %>%
  .[, tax := income_tax(Taxable_Income, "2013-14", .dots.ATO = s1314)] %>%
  .[, .(Taxable_Income, tax)]
```

## `model_income_tax`: modelling changes to personal income tax
While `income_tax` is designed to inflexibly return the tax payable as legislated, 
`model_income_tax` is designed to calculate income tax when changes are made. For example,

```{r model_income_tax}
s1314 %>%
  # reduce top threshold from 180,000 to 150,000
  model_income_tax(ordinary_tax_thresholds = c(0, 18200, 37000, 80000, 
                                               150e3), 
                   baseline_fy = "2013-14") %>%
  .[, .(Taxable_Income, baseline_tax, new_tax)]
```


## `project`
Given a sample file, we can project forward a number of years

```{r s1617}
s1617 <- project(s1314, h = 3L)
```

or to a particular financial year

```{r s1718}
s1718 <- project_to(s1314, "2017-18")
```

Together with `model_income_tax`, this allows us to make point-predictions of future years. The function `revenue_foregone` prettily prints the resultant revenue:

```{r model1819}
sample_file_1314 %>%
  project_to("2018-19") %>%
  model_income_tax(ordinary_tax_thresholds = c(0, 18200, 37000, 87000, 
                                               150e3), 
                   baseline_fy = "2017-18") %>%
  revenue_foregone

```






# NEWS
```{r NEWS, results='asis', echo=FALSE}
cat(readLines("NEWS.md"), sep = "\n")
```

# CRAN Notes
```{r CRAN-comments, results='asis', echo=FALSE}
cat(readLines("cran-comments.md"), sep = "\n")
```



