---
title: "Companion to 2013-14 sample file"
author: "Hugh Parsonage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{companion-to-sample-file-1314}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitrOpts}
options("scipen" = 99)
library(knitr)
opts_chunk$set(fig.width = 9, fig.height = 7)
```

```{r}
FY.YEAR <- "2013-14"
```

```{r loadPackages}
library(data.table)
library(taxstats)
library(grattan)
library(dtplyr)
library(dplyr)
library(ggplot2)
library(scales)
library(magrittr)
library(viridis)
```

```{r grattan_dollar}
grattan_dollar <- function (x, digits = 0) {
  #
  nsmall <- digits
  commaz <- format(abs(x), nsmall = nsmall, trim = TRUE, big.mark = ",", 
                   scientific = FALSE, digits = 1L)
  
  if_else(x < 0, 
          paste0("\U2212","$", commaz),
          paste0("$", commaz))
}
```

```{r load-sample-file}
sample_file <- sample_files_all[fy.year == FY.YEAR]
sample_file <- merge(sample_file, age_range_decoder, by = "age_range")
PREV.FY.YEAR <- yr2fy(fy2yr(FY.YEAR) - 1)
sample_file_prev <- sample_files_all[fy.year == PREV.FY.YEAR]
sample_file_prev <- merge(sample_file_prev, age_range_decoder, by = "age_range")
```

```{r impute-ages}
set.seed(48031)
sample_file %<>%
  group_by(age_range_description) %>%
  mutate(min_age = ifelse(grepl("to", age_range_description), 
                          as.numeric(stringr::str_extract(age_range_description, "^[0-9]{2}")), 
                          ifelse(grepl("70", age_range_description),
                                 70, 
                                 15)),
         max_age = min_age + 5, 
         age_imp = runif(n(), min_age, max_age)) %>%
  select(-min_age, -max_age)
```

```{r tax-bracket}
sample_file %<>%
  mutate(Tax_Bracket = cut(Taxable_Income, 
                           breaks = c(-Inf, 18200, 37e3, 80e3, 180e3, Inf),
                           include.lowest = TRUE, 
                           labels = c("$0-$18,200", 
                                      "$18,201-$37,000", 
                                      "37,001-$80,000", 
                                      "$80,001-$180,000", 
                                      "$180,000+")))
```

```{r texNum}
texNum <- function(number, sig.figs = 3L, dollar = FALSE, pre.phrase = NULL, .suffix = NULL){
  orig.number <- number
  stopifnot(is.numeric(number), length(number) == 1L)
  is.negative <- number < 0
  number <- abs(number)
  if (number == 0){
    warning("Returning 0")
    return(0)
  } else {
    if (is.null(.suffix)){
    n.digits <- ceiling(log10(number))
    
    suffix <- NULL
    suffix_val <- 1
    
    if (n.digits < sig.figs){
      prefix <- signif(x = number, digits = sig.figs)
    } else {
      
      if (n.digits <= 6) {
        prefix_val <- round(number, sig.figs - n.digits - 1)
        prefix <- prettyNum(prefix_val, big.mark = ",", scientific = FALSE)
      } else {
        # Want to show only the number / 10^(multiple of 3) then the suffix multiplier
        suffix_val <- 10 ^ (3 * ((n.digits %/% 3)))
        prefix_val <- signif(number/suffix_val, digits = sig.figs)
        prefix <- prefix_val
        
        if (suffix_val <= 10^12){
          switch(log10(suffix_val) / 3 - 1,
                 suffix <- "~million", 
                 suffix <- "~billion", 
                 suffix <- "~trillion")
        } else {
          prefix <- signif(number / 10^12, digits = sig.figs)
          suffix <- "~trillion"
        }
      }
    }
    } else {
      stopifnot(.suffix %in% c("million", "billion", "trillion"))
      switch(.suffix, 
             "million" = {
              prefix <- signif(number / 10^6, digits = sig.figs)
              suffix <- "~million"
              suffix_val <- 10^6
             }, 
             "billion" = {
               prefix <- signif(number / 10^9, digits = sig.figs)
               suffix <- "~billion"
               suffix_val <- 10^9
             }, 
             "trillion" = {
               prefix <- signif(number / 10^12, digits = sig.figs)
               suffix <- "~trillion"
               suffix_val <- 10^12
             })
      prefix_val <- prefix
    }
    
    if (dollar){
      out <- paste0("\\$", prefix, suffix)
    } else {
      out <- paste0(prefix, suffix)
    }
    
    if (is.negative){
      # General LaTeX
      out <- paste0("\\(-\\)", out)
    }
    # is the displayed number larger than the original?
    if (!is.null(pre.phrase)){
      out_larger <- prefix_val * suffix_val > orig.number
      
        if (out_larger) {
          out <- paste(pre.phrase[1], out, sep = if(grepl("~$", pre.phrase[1])) "" else " ")
        } else {
          if (!isTRUE(all.equal(prefix_val * suffix_val, orig.number, tolerance = .Machine$double.eps)))
            out <- paste(pre.phrase[2], out, sep = if(grepl("~$", pre.phrase[2])) "" else " ")
        }
      
    }
    return(out)
  }
}
```

There were `r texNum(sum(sample_file$WEIGHT), sig.figs = 3L)` taxpayers in `r FY.YEAR` in Australia. Of those, `r texNum(sum(sample_file[Taxable_Income == 0]$WEIGHT))` had zero taxable income (or a taxable loss). (... and so these ''taxpayers'' naturally paid no tax. Nor did the `r texNum(sum(sample_file[Taxable_Income < 18200]$WEIGHT), sig.figs = 2)` individuals below the tax-free threshold. For this report, a *taxpayer* is anyone who lodged a tax return, regardless of their tax liability. 

```{r taxable-income-quantiles}
tx_inc_q <- function(q){
  quantile(sample_file$Taxable_Income, probs = q)
}

my_labs <- grattan_dollar(tx_inc_q((0:10)/10))
my_labs[seq(2, 10, 2)] <- paste0("\n", my_labs[seq(2, 10, 2)])

dens <- density(sample_file[Taxable_Income < tx_inc_q(0.95)]$Taxable_Income)
DF <- with(dens, data.frame(x, y))

sample_file %>%
  mutate(Taxable_Income_decile = ntile(Taxable_Income, 10)) %>%
  filter(between(Taxable_Income, 0, tx_inc_q(0.95))) %>%
  ggplot(aes(x = Taxable_Income)) + 
  geom_density() + 
  scale_fill_viridis(discrete = TRUE) + 
  scale_x_continuous("Taxable Income deciles", 
                     labels = c(my_labs, grattan_dollar(tx_inc_q(0.95))),
                     # limits = c(0, tx_inc_q(0.95)),
                     breaks = c(tx_inc_q((0:10)/10), tx_inc_q(0.95))) + 
  scale_y_continuous(expand = c(0,0)) +
  theme(legend.position = "none", 
        axis.line.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank())
```

```{r taxable-income-quantiles-filled}
DF %>% 
  mutate(Taxable_Income_decile = cut(x, 
                                     breaks = quantile(sample_file$Taxable_Income,
                                                       probs = c(0:10)/10), 
                                     right = TRUE,
                                     include.lowest = TRUE)) %>%
  filter(between(x, -1, tx_inc_q(0.95) * 1.05)) %>%
  {
    ggplot(., aes(x = x, y = y)) + 
      geom_area(color = "black", size = 1.45) +
      geom_area(aes(x = x, y = y, 
                    group = Taxable_Income_decile, 
                    fill = factor(Taxable_Income_decile),
                    color = factor(Taxable_Income_decile))) + 
      scale_color_viridis(discrete = TRUE) + 
      scale_fill_viridis(discrete = TRUE) + 
      scale_x_continuous("Taxable Income deciles", 
                         labels = c(my_labs, grattan_dollar(tx_inc_q(0.95))),
                         expand = c(0,0),
                         # limits = c(-1, tx_inc_q(0.95)*1.05),
                         breaks = c(tx_inc_q((0:10)/10), tx_inc_q(0.95))) + 
      scale_y_continuous(expand = c(0,0), limits = c(0, max(.$y) * 1.05)) +
  theme(legend.position = "none", 
        axis.line.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank())+ 
      
      annotate("text",
               x = tx_inc_q(0.925), 
               y = 2 * max(.$y[.$x > tx_inc_q(0.925)]),
               size = 10/(14/5),
               label = paste0("5% of taxpayers\nhad incomes\ngreater than\n", grattan_dollar(tx_inc_q(0.95))),
               hjust = 0,
               vjust = 0) + 
      annotate("segment", 
               arrow = arrow(type = "closed", length = unit(11, "pt"), angle = 20),
               x = tx_inc_q(0.925), 
               y = 1.9 * max(.$y[.$x > tx_inc_q(0.925)]),
               size = 1,
               xend = tx_inc_q(0.95),
               yend = 1.9 * max(.$y[.$x > tx_inc_q(0.925)])) 
  }
```

