---
title: "Differential uprating: more precise projections of population income"
author: "Hugh Parsonage and William Young"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
# bibliography: bibliography.bib
# csl: vancouver-superscript-brackets-only-year.csl
vignette: >
  %\VignetteIndexEntry{Differential Uprating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, fig.width=7, fig.height=6)
```


Since the most recent versions of data detailing population income and wealth (as is contained within the ATO's sample files) is typically years out-of-date on release, even current situation requires a forecast procedure. 
Since most variables relating to income increase due to inflation, this procedure is called **uprating**.

## Basic uprating

The most basic uprate method is the identity function, where the variables are not changed.

Alternatively, the variables can be increased by a long-term inflation rate. 
Even better, the variables can be increased by the relevant index. 
For instance, salary and wage variables can be increased by the observed increase in the wage price index amd growth in the population of taxpayers by the labour force index.

## Differential uprating

Basic uprating suffers some limitation. 
There are significant structural differences in how much individuals' incomes grow. 

```{r}
library(grattan)
library(data.table)
library(taxstats)
library(dplyr)
library(dtplyr)
library(magrittr)
library(ggplot2)
library(scales)
# library(grid)
# library(gridExtra)
# library(plotly)
```

```{r}
wage_r_by_fy <- 
  data.table(fy.year = yr2fy(2005:2014)) %>%
  mutate(lag_fy = yr2fy(2004:2013)) %>%
  mutate(wage_growth_r = wage_inflator(from_fy = lag_fy, to_fy = fy.year) - 1)
```


```{r, fig.width=7, fig.height=6}
average_salary_by_fy_swtile <- 
  sample_files_all %>%
  select(fy.year, Sw_amt) %>%
  filter(Sw_amt > 0) %>%
  group_by(fy.year) %>%
  mutate(`Salary percentile` = ntile(Sw_amt, 100)) %>%
  ungroup %>%
  group_by(fy.year, `Salary percentile`) %>%
  summarise(average_salary = mean(Sw_amt)) %>%
  ungroup %>%
  arrange(`Salary percentile`, fy.year) %>%
  group_by(`Salary percentile`) %>%
  mutate(r_average_salary = average_salary / lag(average_salary) - 1) %>%
  filter(fy.year != min(fy.year))

{
  p <- 
    average_salary_by_fy_swtile %>%  # NA
    ungroup %>%
    merge(wage_r_by_fy, by = "fy.year") %>%
    mutate(`Basic wage inflator` = "Basic wage inflator") %>%
    ggplot() + 
    geom_area(aes(x = `Salary percentile`, y = r_average_salary, group = fy.year, fill = fy.year), 
              se = FALSE, stat = "smooth", method = "loess") + 
    scale_fill_manual(values = gpal(10)) + 
    theme_bw() + 
    theme(legend.position = "right", plot.background = element_blank()) + 
    geom_line(aes(x = `Salary percentile`, y = wage_growth_r, group = `Basic wage inflator`, color = `Basic wage inflator`),
              size = 1.125) + 
    scale_color_manual(values = "black") +
    scale_y_continuous(name = "Salary rate of increase", label = percent) + 
    facet_wrap(~fy.year, ncol = 5) +
    guides(fill = FALSE) + 
    theme(legend.position = c(0, 1), 
          legend.title = element_blank(), 
          legend.key = element_blank(),
          legend.justification = c(0, 1), 
          legend.margin = unit(0.1, "lines"))
  
  # ggplotly(p)
  p
}

```


```{r}
differential_uprates <- 
  average_salary_by_fy_swtile %>%
  group_by(`Salary percentile`) %>%
  summarise(avg_r = mean(r_average_salary)) %>% 
  mutate(avg_r_normed = avg_r / mean(avg_r))

differential_uprates %>%
  grplot(aes(x = `Salary percentile`, y = avg_r)) +
  geom_line() + 
  scale_y_continuous(label = percent)
```


**Differential uprating** is a method by which variables are uprated by a function of not only the time period and the type of variable, but of the variable's value too. 

```{r}
sw_100tiles <- 
  sample_files_all %>%
  setkey(fy.year, Sw_amt) %>%
  select(fy.year, Sw_amt) %>%
  filter(Sw_amt > 0) %>%
  group_by(fy.year) %>%
  mutate(`Salary percentile` = ntile(Sw_amt, 100)) %>%
  ungroup %>%
  group_by(fy.year, `Salary percentile`) %>%
  summarise(min_sw = min(Sw_amt))
```

```{r}
differentially_uprate_wage_prototype <- function(wage = 1, from_fy, to_fy){
  input <- 
    data.table(fy.year = from_fy, 
               wage = wage) %>%
    mutate(`_order` = 1:n()) %>%
    setkey(fy.year, wage)
  
  merge(differential_uprates, sw_100tiles, by = "Salary percentile") %>%
    mutate(wage = min_sw) %>%
    setkey(fy.year, wage) %>%
    .[input, roll = "nearest"] %>%
    mutate(x = (avg_r_normed * (wage_inflator(from_fy = from_fy, to_fy = to_fy) - 1) + 1) * wage) %>%
    arrange(`_order`) %$%
    x
}

```

## Todos:

1. Account for variation in population year-on-year.




