---
title: "Simple modelling of superannuation changes"
author: "Hugh Parsonage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling superannuation changes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
required_packages <-
  c("testthat", 
    "taxstats",
    "dplyr",
    "rmarkdown", 
    "dtplyr",
    "ggplot2",
    "scales", 
    "broom", 
    "knitr",
    "survey", 
    "viridis",
    "ggrepel")

required_packages_present <-
  all(vapply(required_packages, requireNamespace, quietly = TRUE, FALSE))
knitr::opts_chunk$set(message = FALSE, 
                      eval = required_packages_present)
```


## Introduction
Australia's superannuation system offers a number of tax breaks. Relative to most methods of savings, less tax is paid on money contributed to a super fund, and less tax is paid on the earnings. The functions described here attempt to model changes to tax breaks on money contributed to a super fund.

## Modelling

### The basic functions
The two tax methods modelled are **Superannuation contributions tax concessions** and **Division 293 tax**. The two main outputs are: the extra tax payable by a particular individual (and the symmetric extra revenue), and the number of individuals affected.

To obtain these results (say for 2017-18)

```{r}
library(grattan)
library(hutils)
library(data.table)
if (requireNamespace("taxstats", quietly = TRUE) &&
    packageVersion("taxstats") >= package_version("0.0.5") &&
    !identical(Sys.getenv("NOT_CRAN"), "true")) {
  library(taxstats)
} else {
  templib <- file.path(tempdir(), "model")
  hutils::provide.dir(templib)
  install.packages("taxstats",
                   lib = templib,
                   repos = "https://hughparsonage.github.io/drat/",
                   type = "source")
  library("taxstats", lib.loc = templib)
}
library(magrittr)

sample_files_all <-
    rbindlist(lapply(list(`2003-04` = sample_file_0304, 
                          `2004-05` = sample_file_0405, `2005-06` = sample_file_0506, 
                          `2006-07` = sample_file_0607, `2007-08` = sample_file_0708, 
                          `2008-09` = sample_file_0809, `2009-10` = sample_file_0910, 
                          `2010-11` = sample_file_1011, `2011-12` = sample_file_1112, 
                          `2012-13` = sample_file_1213, `2013-14` = sample_file_1314), 
                     data.table::as.data.table),
              use.names = TRUE,
              fill = TRUE, 
              idcol = "fy.year")
sample_files_all[, WEIGHT := hutils::if_else(fy.year > '2010-11', 50L, 100L)]
#' dollar scales
#' 
#' @name grattan_dollar
#' @param x A numeric vector
#' @param digits Minimum number of digits after the decimal point. (\code{nsmall} in \code{base::format}).
#' @details Makes negative numbers appear as \eqn{-\$10,000} instead of \eqn{\$-10,000} in \code{scales::dollar}.
#' @export
# from scales

grattan_dollar <- function (x, digits = 0) {
  #
  nsmall <- digits
  commaz <- format(abs(x), nsmall = nsmall, trim = TRUE, big.mark = ",", 
                   scientific = FALSE, digits = 1L)
  
  hutils::if_else(x < 0, 
          paste0("\U2212","$", commaz),
          paste0("$", commaz))
}

(new_revenue <- 
  sample_file_1314 %>%
  project_to(to_fy = "2017-18") %>%
  as.data.table %>%
  revenue_from_new_cap_and_div293(new_cap = 25e3, fy.year = "2016-17", new_age_based_cap = FALSE, new_div293_threshold = 250e3))

paste(grattan_dollar(new_revenue / 1e9), "bn")

(n_affected <-
  sample_file_1314 %>%
  project_to(to_fy = "2017-18") %>%
  as.data.table %>%
  n_affected_from_new_cap_and_div293(new_cap = 25e3, fy.year = "2016-17", new_age_based_cap = FALSE, new_div293_threshold = 250e3))

prettyNum(round(n_affected), big.mark = ",")
```

Notes: 

  1. `fy.year` refers to the year the function takes the tax scales from, not the forecast year. Because we don't have tax scales for 2017-18 yet, we model using the most recent (`2017-18`).
  2. The functions require their inputs to be `data.table`s. Sorry if you don't like `data.table`s.
  3. You must use `sample_file_1314` (because they have superannuation contributions variables).

### Distributional analysis

#### By taxable income decile
Let's create an object for `sample_file_1718` avoid recreating it every time:
```{r}
sample_file_1718 <-
  sample_file_1314 %>%
  project_to(to_fy = "2017-18") %>%
  as.data.table
```

The functions mentioned earlier return single values. In contrast, `model_new_caps_and_div293` returns the sample file with extra variables, which can then be analyzed as a standard sample file. The variables of note are `prv_revenue` which is the tax payable under the old system, and `new_revenue` which is the tax payable under the proposed system. Thus:

```{r}
new_sample_file_1718 <- 
  sample_file_1718 %>%
  model_new_caps_and_div293(new_cap = 25e3, fy.year = "2016-17", new_age_based_cap = FALSE, new_div293_threshold = 250e3)

```

```{r}
library(knitr)
library(dplyr)
library(dtplyr)  # for data.table

new_sample_file_1718 %>%
  mutate(Taxable_Income_decile = ntile(Taxable_Income, 10)) %>%
  group_by(Taxable_Income_decile) %>%
  summarise(`Average increase in tax` = round(mean(new_revenue - prv_revenue), 2)) %>%
  arrange(Taxable_Income_decile) %>%
  kable
```

```{r}
library(ggplot2)
new_sample_file_1718 %>%
  mutate(Taxable_Income_decile = ntile(Taxable_Income, 10)) %>%
  group_by(Taxable_Income_decile) %>%
  summarise(`Average increase in tax` = mean(new_revenue - prv_revenue)) %>%
  arrange(Taxable_Income_decile) %>%
  #
  mutate(`Taxable income decile` = factor(Taxable_Income_decile)) %>%
  ggplot(aes(x = `Taxable income decile`, y = `Average increase in tax`)) + 
  geom_bar(stat = "identity") + 
  
  # cosmetic:
  scale_y_continuous(label = grattan_dollar) + 
  theme(axis.title.y = element_text(face = "bold", angle = 90, margin = margin(1, 1, 1, 1, "lines")))
```






